Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-04 paper.tex
Added section on drug target networks. Updated bib entries

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined,
it has become apparent that this model does not always match
reality. Instead, it has become increasingly clear that many drugs
(and more generally, small molecules) exhibit polypharmacological
behavior - interacting with multiple targets, including their putative
primary target. In many cases, the activity against the primary target
is significantly higher than off-target activities, minimizing the
effects of the latter. However, this is not always the case as in the
example of terfenadine (a histamine H1 antagonist), where off target
activity against hERG led to an increased risk of cardiac death. But
more importantly, it is becoming increasingly clear that many diseases
are not driven by dysfunction of a single a target. Examples include
diabetes and various cancers. In parallel, even when a single target
is the main driver for a given disease state, compensatory pathways
can ``route around'' inhibition of the primary target, thus restoring
the functional behavior that the primary target is responsible (i.e.,
the system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. Finally, it should be noted
that while the preceding discussion presents polypharmacological
behavior as a negative characteristic, this is not always the
case. Indeed, modern approaches to drug development now consider such
behavior attractive leading to efforts to rationally design or control
polypharmacology \cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

Given these observations it is clear that pathways play fundamental
roles in biological systems. Naturally, approaches that analyse these
networked systems can be effective in understanding their behavior and
then identifying ways to modulate them. The remainder of this chapter
looks at how network methods can be applied to various topics in drug
discovery ranging from drug-target interactions to analyzing
structure-activity relationships.

@@

Network analysis has also found its application in Medicinal Chemistry and High-throughput Screening. Cheminformatics based chemical scaffold and topology networks capture the structure-activity relationships between compounds in screening experiments \cite{varin2011mining}. These help in medicinal chemistry efforts towards the identification and optimization of lead compounds for the  biological target of interest by allowing visualization of the chemical space in the backdrop of biological activity.

@@

>>>>

The traditional, reductionist view of biology has considered the
behavior and interactions of individual molecular entities as a way to
understand and explain biological phenomena. This approach has become
embodied in the ``one-target, one-drug'' approach to drug
development. Leaving aside the subtleties of how targets are defined
\cite{Imming:2006lt}, it has become apparent that this model does not
always match reality. Instead, it has become increasingly clear that
many drugs (and more generally, small molecules) exhibit
polypharmacological behavior - interacting with multiple targets,
including their putative primary target. In many cases, the activity
against the primary target is significantly higher than off-target
activities, minimizing the effects of the latter. However, this is not
always the case as in the example of terfenadine (a histamine H1
antagonist), where off target activity against hERG
\cite{Crumb:1995it} led to an increased risk of cardiac death. But
more importantly, it is becoming increasingly clear that many diseases
are not driven by dysfunction of a single a target. Examples include
diabetes and cancer. In parallel, even when a single target is the
main driver for a given disease state, compensatory pathways can
``route around'' inhibition of the primary target, thus restoring the
functional behavior that the primary target is responsible (i.e., the
system is now resistant to the original drug)
\cite{Sierra:2010mz,Dienstmann:2012qr}. Finally, it should be noted
that while the preceding discussion presents polypharmacological
behavior as a negative characteristic, this is not always the
case. Indeed, modern approaches to drug development now consider such
behavior attractive leading to efforts to rationally design or control
polypharmacology \cite{Gujral:2014dp,Ciceri:2014fv,Metz:2010ys}.

Given these observations it is clear that pathways play fundamental
roles in biological systems. Naturally, approaches that analyse these
networked systems can be effective in understanding their behavior and
then identifying ways to modulate them. Importantly, network
approaches provide the opportunity to integrate multiple entity types
(small molecules, protein targets, genes) in a systematic fashion.The
remainder of this chapter looks at how network methods can be applied
to various topics in drug discovery ranging from drug-target
interactions to analyzing structure-activity relationships.

@@

Network analysis has also found its application in medicinal chemistry
and high-throughput screening. Cheminformatics based chemical scaffold
and topology networks capture the structure-activity relationships
between compounds in screening experiments \cite{Varin:2011ve}. These
help in medicinal chemistry efforts towards the identification and
optimization of lead compounds for the biological target of interest
by allowing visualization of the chemical space in the backdrop of
biological activity.

@@

Networks constructed from drugs and their protein targets were first
described by Yildirim et al \cite{Yildirim:2007aa}, who considered
approved and investigational drugs. Networks built using the former
subset highlighted a number of ``follow on'' drugs (i.e., drugs
targeting a protein that is already targeted by another drug. When
investigational drugs were included in the analysis, the resultant
network highlighted a more functionally diverse set of protein
targets. Importantly, the drug-target network representation
effectively highlights the polypharmacology exhibited by many
drugs. They also constructed a drug-target protein network by defining
protein targets as nodes and connecting two nodes if they are targeted
by the same drug. But integrating this network with a protein-protein
interaction network they were able to highlight the fact that drug
target proteins were not as essential \cite{Jeong:2001gd} as high
degree PPI nodes. While many drug-target networks have been
constructed from experimentally derived or manually curated data, a
number of methods to predicting drug-target interactions have also
been described
\cite{Alaimo:2013ev,Wang:2013mw,Heiskanen:2013rc}. However, it is
still an open question as to whether such predicted networks can serve
as the foundation for downstream analyses, though Pahikkala et al
\cite{Pahikkala:2014gt} describe a number of guidelines to enhance the
reliability of predicted drug-target interactions.

Many workers have described analytic approaches based on drug-target
networks. For example, the MetaCore and MetaDrug suite of tools have
been used to perform a variety of functional analyses on small
molecules, particularly metabolites \cite{Brennan:2009kx}.  Examples
include pathway enrichment analyses and pathway reconstruction. The
latter is useful in placing new or predicted metabolites in the
context of known protein-protein and protein-compound
interactions. Iwata et al \cite{Iwata:2013lk} described the use of a
drug-target network to identify protein domains that are associated
with drug side effects. Wang et al \cite{Wang:2013gn} characterized
the topological properties of drug-target networks and noted that
target essentiality and centrality were strongly linked to drug side
effects and that a drugs' side effects were primarily influenced by
the number of \emph{essential} targets rather than the number of drug
targets. Drug-target networks have been used as input to a variety of
predictive models. For example, a model to predict ATC codes from a
drug-target network was described by Wang et al
\cite{Wang:2013ye}. Tabei et al \cite{Tabei:2012tn} used a drug-target
network a input to a classifier that was used to extract chemogenomic
features. These features represent associations between chemical
substructures and protein domains. The set of features that provides a
compressed representation of the drug-target network and the authors
proposed that they could be used to suggest new substructures for a
given protein domain.

A number of workers have used drug-target network information as the
basis of drug repositioning methodologies. Wang et al
\cite{Wang:2014lp} who combined drug-target and disease-target
networks to generate repositioning hypotheses. Tan et al
\cite{Tan:2014qb} employed drug-target networks as a component of a
larger workflow that included chemical structure similarity, gene
semantic similarity to identify new indications for a number of
preexisting drugs.  Peng et al \cite{Peng:2014vy} considered a variant
of the drug-target network concept by developing an interaction
network of FDA approved drugs and protein pockets. Drugs were docked
into each pocket and the docking score was used to define edges
between the drug and protein whose pocket it was docked to.
